Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-09 paper.qmd
model removwed
Added UMAP layout for S-curve data

We define the NLDR model as function $g\text{:}~ \mathbb{R}^p \rightarrow \mathbb{R}^k$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that maps the entire space $\mathbb{R}^p$ to a subset of $\mathbb{R}^k$ is suitable. <!--Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.-->

@@

>>>>

We define the NLDR as a function $g\text{:}~ \mathbb{R}^p \rightarrow \mathbb{R}^k$, with (hyper-)parameters $\mathbfit{\theta}$. The parameters, $\mathbfit{\theta}$, depend on the choice of $g$, and can be considered part of model fitting in the traditional sense. Common choices for $g$ include functions used in tSNE, UMAP, PHATE, TriMAP, PaCMAP, or MDS, although in theory any function that maps the entire space $\mathbb{R}^p$ to a subset of $\mathbb{R}^k$ is suitable. <!--Any input requirements for the data (such as normalization, or preprocessing through the use of PCA or similar) is considered part of the function $g$.-->

@@

<!--UMAP applied for Scurve data-->
```{r}
umap_scurve <- read_rds(file = "data/s_curve/s_curve_umap.rds")

nldr_scurve <- umap_scurve |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) +
  interior_annotation("a")
```

<!--Full hexagonal grid with UMAP data-->

<!--Non-empty bins with bin centroids-->

<!--2D model-->

```{r}
#| label: fig-NLDR-scurve
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Researchers may have seen any of these in their analysis of this data, depending on their choice of method, or typical parameter choice. Would they make different decisions downstream in the analysis depending on which version seen? Which is the most accurate representation of the structure in high dimensions?"
#| fig-width: 8
#| fig-height: 4
#| out-width: 100%

nldr_scurve +
  plot_layout(ncol = 4)
```
